---
title: "Final Project"
author: 'Peter Lehrman'
subtitle: 'due Monday March 28 11:59 PM'
output:
  html_document:
    df_print: paged
  pdf_document: default
editor_options:
  chunk_output_type: console
---

```{r include = FALSE}
knitr::opts_chunk$set(warning = FALSE)
```


```{r}
library(tidyverse)
library(fivethirtyeight)
```
 approve your first choice -- NBA ELO data.

Good start at a research question. Moving forward, you'll want to think about how you will quantify playoff performance. E.g. total points scored? Difference between opponent and home points? Total number of wins? etc.

You might think of other creative extensions to incorporate some of additional variables like game_result

e.g.

- how does ELO score correlate with probability of an away win? probability of a home win? Does ELO ranking align sort franchises according to some other metric, e.g. total # of wins in a season, etc?
This could be useful for identifying confounding.

```{r load-data, message = FALSE}
nba_data <- read_csv("data/nbaallelo.csv")
```